### 2023-02-26 calculus.tex
Further describe uexp and mexp typing rules

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. Both operands of an addition expression
$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
$\ETrue$ synthesizes $\TBool$, which is inconsistent with $\TNum$.

@@

The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
This comes as a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} ($\EMName$), which compose the surface-level syntax. Types
    classify unmarked expressions according to a standard bidirectional gradual typing semantics.

  \item \textbf{Marked expressions} ($\ECMName$), which mirror the structure of unmarked expressions
    but are extended with \textbf{error markings}. Every marked expression corresponds to an
    unmarked expression by \emph{mark erasure}.

  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. While the semantics that
follow concern a gradually typed language, a non-gradual system may be recovered straightforwardly
by removing the unknown type and replacing type consistency with equality.

In the interest of space, this overview focuses on the ``interesting'' cases. All rules and
metatheorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant (discussed in \cref{sec:calculus-agda}).

@@

Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
$\EMV$ synthesizes or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected. The \emph{subsumption} rule (ASubsume) states that if an expression
synthesizes a type, it may be analysed against that type or any \emph{consistent} type.

Note that the subsumption rule requires that $\EMV$ be ``subsumable'', a judgment defined on all
forms except lambda abstractions and if-else expressions. This restricts the usage of subsumption to
a rule of ``last resort'', admitting the unicity of typing.

@@

The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
additionally consistent with any other type, and arrow types are consistent if their components are.

$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$ gives that $\TMV$ is a \emph{matched arrow type}, which means
that it may be considered an arrow type. It is purely a technical mechanism to avoid duplication of
rules related to arrow types.

@@

@@

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesized type
    of $\ECMV$ is inconsistent with the required type.

@@

Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
the rules for corresponding unmarked forms. \Cref{fig:calculus-typing-mexp} contains a selection of
the ``interesting'' cases, particularly those concerned with error markings.

@@

For unbound variables $\ECUnbound{x}$, nothing may be said about their types; hence, they synthesize
the unknown type. Inconsistent branch forms behave similarly, as inconsistent types have no join.

There are three rules concerned with the typing of lambda abstractions. MALam1, applies to
$\ECLam{x}{\TMV}{\ECMV}$ if the type $\TMV_3$ analysed against is a matched arrow type
$\TArrow{\TMV_1}{\TMV_2}$ and input type $\TMV_1$ is consistent with the binding ascription $\TMV$.
This corresponds to the lambda rule for unmarked expressions. However, we are additionally concerned
with the cases when $\TMV_3$ is not a matched arrow type or $\TMV_1$ is inconsistent with $\TMV$.
These are described by MALam2 and MALam3.

Alongside the standard subsumption rule, in which an expression that synthesizes a type may be
analysed against any consistent type, MAInconsistentTypes governs most occurrences of the
inconsistent types marking: $\ECInconType{\ECMV}$ analyzes against any type as long as $\ECMV$
synthesizes an inconsistent type. In both cases, we stipulate that subsumption may be applied to
$\ECMV$, written $\subsumable{\ECMV}$. In particular, lambda abstractions and if-else forms are
\emph{not} subsumable, formalizing these rules as ones of ``last resort'', preserving determinism of
typing.

Similarly to lambda abstractions, two rules govern application. In the unmarked syntax, the
application $\EAp{\EMV_1}{\EMV_2}$ was well-typed if the type $\TMV$ synthesized by $\EMV_1$ was a
matched arrow type. This corresponds to the MSAp1 rule for marked expressions. In the case that
$\TMV$ is not a matched arrow type, we require an inconsistent types marking around the

\subsection{Mark Insertion}

@@

\subsection{Alternative Formulations}

\subsection{Non-Gradual System}
\label{sec:calculus-non-gradual}

>>>>

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. Both operands of an addition expression
$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
$\ETrue$ synthesises $\TBool$, which is inconsistent with $\TNum$.

@@

The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
This comes as a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} (\cref{sec:calculus-uexp}), which compose the surface-level
    syntax. Types classify unmarked expressions according to a standard bidirectional gradual typing
    semantics.

  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
    unmarked expressions but are extended with \textbf{error markings}. Every marked expression
    corresponds to an unmarked expression by \emph{mark erasure}.

  \item \textbf{Mark insertion} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
semantics that follow concern a gradually typed language, a non-gradual system may be recovered
straightforwardly by removing the unknown type and replacing type consistency with equality.

In \cref{sec:calculus-correctness}, we discuss the metatheory that ensures the system---and any such
error localization semantics---to be correct. We follow then by extending the calculus with support
for product types. In the interest of space, this overview focuses on ``interesting'' cases. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization
(including the extension with products) in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}.

@@

Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
$\EMV$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected.

The \emph{subsumption} rule (ASubsume) states that if an expression synthesises a type, it may be
analysed against that type or any \emph{consistent} type. To invoke subsumption, the expression
$\EMV$ must be ``subsumable'', written $\subsumable{\EMV}$, a judgment defined on all forms except
lambda abstractions and if-else expressions. This restricts the usage of subsumption to a rule of
``last resort'', admitting the unicity of typing.

@@

The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
additionally consistent with any other type, and arrow types are consistent if their components are.
The USAp and UALam rules for application and lambda abstractions are nearly standard, but employ the
judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which gives that $\TMV$ is a \emph{matched arrow
type}, i.e. it may be considered an arrow type. It is purely a technical mechanism to avoid
duplication of rules related to arrow types.

@@

The branches of if-else expressions may synthesize different types $\TMV_1$ and $\TMV_2$ as long as
they are consistent. The entire expression then synthesizes the join, or lowest upper bound,
$\TJoin{\TMV_1}{\TMV_2}$.

@@

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesised type
    of $\ECMV$ is inconsistent with the required type.

@@

Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
the rules for corresponding unmarked forms. Additional rules handle the cases when certain premises
are not met, requiring then that expressions be wrapped in appropriate error marks.
\Cref{fig:calculus-typing-mexp} contains a selection of these ``interesting'' cases.

In parallel with the rule governing bound variables, MSUnbound types unbound variables. Such
variables must be marked unbound, and since nothing may be said about their types, they synthesise
the unknown type. MSInconsistentBranches corresponds to the USIf, handling the case that the
branches synthesise inconsistent types and hence have no join. Such forms are required to be marked
as such.

@@

In addition to a synthesis rule for lambda abstractions that mirrors USLam, observe that there are
three, instead of one, analytic lambda rules. MALam1 applies to $\ECLam{x}{\TMV}{\ECMV}$ if the type
$\TMV_3$ analysed against is a matched arrow type $\TArrow{\TMV_1}{\TMV_2}$ and the input type
$\TMV_1$ is consistent with the binding ascription $\TMV$. This corresponds to UALam.

However, we are additionally concerned with the cases when $\TMV_3$ is not a matched arrow type or
$\TMV_1$ is inconsistent with $\TMV$. These are described by MALam2 and MALam3. The absence of a
matched arrow type means that $\TMV_3$ must be inconsistent the type synthesised by the lambda, so
the former appropriately requires that the lambda be inside an inconsistent types mark. In this
case, as no output type is known, the body is analysed against the unknown type. Similarly,
inconsistency between the matched input type and binding ascription also implies that $\TMV_3$ is
inconsistent with the synthesised type; MALam3 also requires that the lambda be marked inconsistent,
and the body is analysed against $\TMV_2$.

% TODO mention that every expression synthesises?

Two synthesis rules govern application. In the unmarked language, the application
$\EAp{\EMV_1}{\EMV_2}$ is well-typed if the type $\TMV$ synthesised by $\EMV_1$ is a matched arrow
type $\TArrow{\TMV_1}{\TMV_2}$ and the argument $\EMV_2$ analyses against $\TMV_1$. This corresponds
to the MSAp1 rule for marked expressions. In the case that $\TMV$ is not a matched arrow type, i.e.
it may not be applied, we require $\ECMV_1$ to be inside an inconsistent types mark. In this way, the
blame is placed upon $\ECMV_1$, and this approach is in accordance with the meaning of the
inconsistent types mark.

It is worth noting, however, that $\ECMV_1$ is not actually in analytic position but in synthetic
position. Treating such ``constrained synthetic'' mode specially, we may instead designate a
specialized mark, written $\ECInconMatchedArrow{\ECMV_1}$, for such applications. This yields the
following rule:
\[%
  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MSAp2'}
\]%
It then becomes natural to extend this to other elimination forms that require the handling of
unmatched types, such as products. We demonstrate such an extension in \cref{sec:calculus-products}.

Finally, alongside the standard subsumption rule, MAInconsistentTypes governs most occurrences of
the inconsistent types mark $\ECInconType{\ECMV}$ analyses against any type as long as $\ECMV$
synthesises an inconsistent type. In both cases, we again stipulate that subsumption may be applied
to $\ECMV$, written $\subsumable{\ECMV}$, for similar reasons as before.

\subsection{Mark Insertion}
\label{sec:calculus-marking}

@@

% Well-typed expressions have no resulting marks

\subsection{Adding Products}
\label{sec:calculus-products}
